---
title: Plotting consumer and producer surpluses in ggplot2
author: EP
date: '2017-09-19'
image: "/post/2017-09-19-plotting-consumer-and-producer-surpluses-in-ggplot2_files/figure-html/finished_chart-1"
slug: plotting-consumer-and-producer-surpluses-in-ggplot2
categories:
  - R
  - economics
tags: []
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, 
                      fig.height = 3.5, fig.width = 3.5,
                      fig.align = "center")
```

[Andrew Heiss](https://twitter.com/andrewheiss) has an interesting [recent 
post](https://www.andrewheiss.com/blog/2017/09/15/create-supply-and-demand-economics-curves-with-ggplot2/)
where he uses ggplot2 to plot a nicely annotated supply-demand 
diagram. In this post we will take a somewhat different approach to solve the
same problem, with the added feature of filling in the areas representing
consumer and producer surpluses.

We start by defining supply and demand functions.

```{r demand_supply}
library(tidyverse)

demand <- function(q) (q - 10)^2
supply <- function(q) q^2 + 2*q + 8
```

We plot these over a specified domain using `stat_function`.

```{r basic_chart, fig.width = 5}
x <- 0:5

chart <- ggplot() +
  stat_function(aes(x, color = "Demand"), fun = demand) +
  stat_function(aes(x, color = "Supply"), fun = supply)
chart
```

We now need to find the point of intersection, or, in economic terms, the 
equilibrium price and quantity. We use the `uniroot` function for this and pass 
it an anonymous function that calculates the difference between the supply and 
demand. `uniroot` then finds where this difference is zero. This gives us the
equilibrium quantity $q^*$. Passing this quantity to the supply function then
gives us the equilibrium price $p^*$.

```{r equilirium_point}
# Equilibrium quantity
q <- uniroot(function(x) demand(x) - supply(x), range(x))$root

# Equilibrium price
p <- supply(q)
```

We can now annotate the chart with the equilibrium point.

```{r chart_w_equilibrium, fig.width = 5}

chart <- chart + annotate("point", x = q, y = p, color = "grey30")
chart
```

Next we add some dashed segments from the equilibrium point to the axes. 

```{r chart_w_dashed_lines, fig.width = 5}
chart <- chart + 
  annotate("segment", x = q, xend = q, y = 0, yend = p,
           linetype = "dashed", color = "grey30") +
  annotate("segment", x = 0, xend = q, y = p, yend = p,
           linetype = "dashed", color = "grey30")
chart
```

Finally, we want to color the area between the demand curve and the dashed line
as consumer surplus, and the area below the dashed line but above the supply
curve as producer surplus. For this we need to pre-calculate a dummy series from
0 to the equilibrium point, which we pass to the `geom_ribbon` function.

```{r chart_w_ribbon, fig.width = 5}
z <- seq(0, q, 0.01)

chart <- chart + 
  geom_ribbon(aes(x = z, ymin = supply(z), ymax = p,
                  fill = "Producer surplus"), alpha = 0.25) +
  geom_ribbon(aes(x = z, ymin = p, ymax = demand(z),
                  fill = "Consumer surplus"), alpha = 0.25)
chart
```

As a last touch-up we drop some chart junk and properly annotate the equilibrium
price and quantity.

```{r finished_chart}
chart + 
  scale_x_continuous(expand = c(0, 0), 
                     breaks = q, labels = "q*") +
  scale_y_continuous(expand = c(0, 0), 
                     breaks = p, labels = "p*") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = c(1, 1), 
        legend.justification = c(1, 1),
        legend.spacing = unit(0, "cm"), 
        legend.margin = margin(0, 0, 0, 0, "cm")) +
  labs(x = "Quantity", y = "Price", 
       color = NULL, fill = NULL)
```

Indeed, we can now wrap all that code in a function that takes as input a supply
and a demand function, and returns the corresponding diagram.

```{r plot_function}
plot_surpluses <- function(demand, supply, domain) {
  
  # Equilibrium quantity
  q <- uniroot(function(x) demand(x) - supply(x), domain)$root
  
  # Equilibrium price
  p <- supply(q)
  
  # Domain
  x <- seq(domain[1], domain[2], 0.1)
  
  # Dummy domain for geom_ribbon
  z <- seq(0, q, 0.01)
  
  ggplot() +
    stat_function(aes(x, color = "Demand"), fun = demand) +
    stat_function(aes(x, color = "Supply"), fun = supply) +
    geom_ribbon(aes(x = z, ymin = supply(z), ymax = p,
                    fill = "Producer surplus"), alpha = 0.25) +
    geom_ribbon(aes(x = z, ymin = p, ymax = demand(z),
                    fill = "Consumer surplus"), alpha = 0.25) +
    annotate("segment", x = q, xend = q,
             y = 0, yend = p,
             linetype = "dashed", color = "grey30") +
    annotate("segment", x = 0, xend = q,
             y = p, yend = p,
             linetype = "dashed", color = "grey30") +
    annotate("point", x = q, y = p, color = "grey30") +
    scale_x_continuous(expand = c(0, 0), 
                       breaks = q, labels = "q*") +
    scale_y_continuous(expand = c(0, 0), 
                       breaks = p, labels = "p*") +
    theme_minimal() +
    theme(panel.grid = element_blank(),
          legend.position = c(1, 1), 
          legend.justification = c(1, 1),
          legend.spacing = unit(0, "cm"), 
          legend.margin = margin(0, 0, 0, 0, "cm")) +
    labs(x = "Quantity", y = "Price", 
         color = NULL, fill = NULL)
}

plot_surpluses(demand, supply, domain = c(0, 5))
```
