---
title: Drawing an annotated unit circle with ggplot2
author: EP
date: '2017-08-07'
categories:
  - R
  - mathematics
tags:
  - trigonometry
  - ggplot2
---

```{r, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                      fig.width=5, fig.height=5, fig.align='center')
```

*The unit circle: everybody's favorite circle.*

I recently needed to an annotated unit circle for some teaching material I was
preparing. Rather than using one of the countless pictures already available, I
thought it was a good excuse to play around a bit with using mathematical
annotations in `ggplot2`. This post explains the process. 

Here's what we'll be working towards:

```{r, echo=FALSE}
library(tidyverse)

get_circle_coords <- function(r = 1, ...) {
  data_frame(theta = seq(0, 2 * pi, ...),
             x     = cos(theta) * r,
             y     = sin(theta) * r)
}

coords_pi <- get_circle_coords(by = pi / 6) %>% head(-1)
coords_pi$angle <- seq(0, 330, 30) %>% paste(" * degree")
coords_pi$pi <- c(
  "0", "frac(pi, 6)", "frac(pi, 3)", "frac(pi, 2)", 
  "frac(2 * pi, 3)", "frac(5 * pi, 6)", "pi",
  "frac(7 * pi, 6)", "frac(4 * pi, 3)", "frac(2 * pi, 3)", 
  "frac(5 * pi, 3)", "frac(11 * pi, 6)"
)

coords_pi$trig <- c(
  "(1*','* 0)",
  "(frac(sqrt(3), 2) *','* ~ frac(1,2))",
  "(frac(1, 2) *','* ~ frac(sqrt(3), 2))",
  "(0*','* 1)",
  "(-frac(sqrt(3), 2) *','* ~ frac(1,2))",
  "(-frac(1, 2) *','* ~ frac(sqrt(3), 2))",
  "(-1*','* 0)",
  "(-frac(sqrt(3), 2) *','* ~ -frac(1,2))",
  "(-frac(1, 2) *','* ~ -frac(sqrt(3), 2))",
  "(0*','* -1)",
  "(frac(sqrt(3), 2) *','* ~ -frac(1,2))",
  "(frac(1, 2) *','* ~ -frac(sqrt(3), 2))"
)

geom_l <- partial(geom_label, size = 2.5, 
                  label.size = NA, parse = TRUE, color = "firebrick")

ggplot(coords_pi, aes(x = x, y = y)) +
  geom_path(data = get_circle_coords(length.out = 200), color = "grey50") +
  geom_point(color = "grey50") +
  geom_segment(aes(xend = 0, yend = 0), color = "grey50") +
  geom_l(aes(label = angle, x = x / 2, y = y / 2)) +
  geom_l(aes(label = pi, x = x * 4/5, y = y * 4/5)) +
  geom_l(aes(label = trig), fill = NA, 
         vjust = "outward", hjust = "outward") +
  xlim(-1.2, 1.2) +
  ylim(-1.2, 1.2) +
  theme_void() +
  coord_equal()
```

We start by defining a function that, given a radius, generates a data frame of
coordinates of a circle centered around the origin.

```{r}
library(tidyverse)
library(scales)
library(stringr)

get_circle_coords <- function(r = 1, ...) {
  data_frame(theta = seq(0, 2 * pi, ...),
             x     = cos(theta) * r,
             y     = sin(theta) * r)
}

circ <- get_circle_coords(length.out = 200)
qplot(circ$x, circ$y, geom = "path")
```

Next, we want to generate the coordinates where we go around the unit circle by
one-sixth $\pi$ at each step (we drop the last observation to avoid overlap):

```{r}
coords_pi <- get_circle_coords(by = pi / 6) %>% head(-1)
```

We can now plot the circle itself as a `geom_path`, the hubs as `geom_point`,
and spokes as `geom_segment`. `theme_void` drops all unnecessary chart junk, and
`coord_equal` makes sure that one unit on the x-axis is equivalent to one unit
on the y-axis.

```{r}
ggplot(coords_pi, aes(x = x, y = y)) +
  geom_path(data = circ, color = "grey50") +
  geom_point(color = "grey50") +
  geom_segment(aes(xend = 0, yend = 0), color = "grey50") +
  xlim(-1.2, 1.2) +
  ylim(-1.2, 1.2) +
  theme_void() +
  coord_equal()
```

We add a character variable that gives the equivalent angles in degrees:

```{r}
coords_pi$angle <- seq(0, 330, 30) %>% paste(" * degree")
```

To properly typeset the fractions of $\pi$ we use R's built-in support for
mathematical annotation (`?grDevices::plotmath`). Since we only need a few
annotations, we can hard-code these:

```{r}
coords_pi$pi <- c(
  "0", "frac(pi, 6)", "frac(pi, 3)", "frac(pi, 2)", 
  "frac(2 * pi, 3)", "frac(5 * pi, 6)", "pi",
  "frac(7 * pi, 6)", "frac(4 * pi, 3)", "frac(2 * pi, 3)", 
  "frac(5 * pi, 3)", "frac(11 * pi, 6)"
)
```

Finally, we use some more `plotmath` to typeset the coordinates of some of the
extact trigonometric constants. This might look a bit fiddly, but we only need
to figure out the pattern for the first quadrant, and then just make sure we get
the signs right for the other quadrants.

```{r}
coords_pi$trig <- c(
  "(1*','* 0)",
  "(frac(sqrt(3), 2) *','* ~ frac(1,2))",
  "(frac(1, 2) *','* ~ frac(sqrt(3), 2))",
  "(0*','* 1)",
  "(-frac(1, 2) *','* ~ frac(sqrt(3), 2))",
  "(-frac(sqrt(3), 2) *','* ~ frac(1,2))",
  "(-1*','* 0)",
  "(-frac(sqrt(3), 2) *','* ~ -frac(1,2))",
  "(-frac(1, 2) *','* ~ -frac(sqrt(3), 2))",
  "(0*','* -1)",
  "(frac(1, 2) *','* ~ -frac(sqrt(3), 2))",
  "(frac(sqrt(3), 2) *','* ~ -frac(1,2))"
)
```

Since we're going to plot several layers of `geom_label` we can use
`purrr::partial` to partially apply all the common arguments that these will
take:

```{r}
geom_l <- partial(geom_label, size = 2.5, 
                  label.size = NA, parse = TRUE, 
                  color = "firebrick")
```

Lastly, we plot the final illustration:

```{r}
ggplot(coords_pi, aes(x = x, y = y)) +
  geom_path(data = circ, color = "grey50") +
  geom_point(color = "grey50") +
  geom_segment(aes(xend = 0, yend = 0), color = "grey50") +
  geom_l(aes(label = angle, x = x / 2, y = y / 2)) +
  geom_l(aes(label = pi, x = x * 4/5, y = y * 4/5)) +
  geom_l(aes(label = trig), fill = NA, 
         vjust = "outward", hjust = "outward") +
  xlim(-1.2, 1.2) +
  ylim(-1.2, 1.2) +
  theme_void() +
  coord_equal()
```
