---
title: Most popular ggplot2 geoms
author: EP
date: '2017-08-10'
categories:
  - R
tags:
  - ggplot2
---

In this post we will use the `gh` package to search Github for uses of the
different geoms used in `ggplot2`.

We load some necessary packages.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(scales)
library(stringr)
library(gh)
```

Next, we find all the functions exported by `ggplot2`, keep only those of the
form `geom_*`, and put them in a data frame.

```{r geom_names}
geoms <- getNamespaceExports("ggplot2") %>% 
  keep(str_detect, pattern = "^geom") %>% 
  data_frame(geom = .)

geoms
```

Since there are 44 different geoms we will have to make 44 separate calls to the
Github API. To avoid running into a rate limit, we thus define a function
operator called `delay`, which takes as input a function `f` and a number
`delay`, and returns that same function, but now modified so that its call is
delayed by `delay` seconds.

```{r func_ops}
delay <- function(f, delay = 3) {
  function(...) {
    Sys.sleep(delay)
    f(...)
  }
}
```

When web scraping and making API calls it's often nice to get some real-time
feedback on what calls are being made. For this we define a second function
operator, `w_msg`, which modifies a function so that it prints out its first
argument whenever it is called.

```{r}
w_msg <- function(f) {
  function(...) {
    args <- list(...)
    message("Processing: ", args[[1]])
    f(...)
  }
}
```

Our workhorse function will be `search_gh`, which takes a search query parameter
`q`, and then searches Github for that query. (NB: if you want to run this
yourself, you should follow the instructions in `?gh::gh_whoami` for how to
handle your Github API token).

```{r}
search_gh <- function(q, ...) {
  gh("/search/code", q = q, ...)
}
```

With all this in place, we simply map our modified `search_gh` over all 44
geoms, and extract the `total_count` field.

```{r, eval=TRUE, cache=TRUE}
geoms <- geoms %>% 
  mutate(
    result      = map(geom, w_msg(delay(search_gh))),
    total_count = map_int(result, "total_count")
)
```

```{r}
geoms
```

Finally, we plot the results.

```{r, fig.width=4, fig.height=6, fig.align='center'}
ggplot(geoms, aes(x = total_count, 
                  y = reorder(geom, total_count))) +
  geom_point(color = "red") +
  geom_text(
    aes(label = str_replace(geom, "geom_(.*)", "\\1  ")),
    size = 3,
    hjust = 1,
    color = "grey30"
  ) +
  scale_y_discrete(expand = c(0.03, 0)) +
  scale_x_log10(
    limits = c(100, 1500000),
    expand = c(0, 0),
    breaks = 10 ^ c(1:6),
    labels = format(10 ^ c(1:6),
                    scientific = FALSE, big.mark = ",")
  ) +
  annotation_logticks(sides = "b") +
  theme_minimal() +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.x = element_blank(),
    plot.margin = unit(rep(5, 4), "mm")
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Make geom_qq Great Again",
    subtitle = paste0("Most popular ggplot2 geoms,", 
                      " by total count on Github")
  )
```
