---
title: Probability kata
author: EP
date: '2017-12-20'
categories:
  - R
  - mathematics
tags:
  - probability
  - monet carlo
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
set.seed(2)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### Introduction

While having dinner with a colleague last night, my colleague told me that he'd
played a dice game with a risk-loving opponent (who also apparently made a lot
of money on the Bitcoin bubble). My colleague said that they had each rolled a
six-sided die ten times, and that his friend had (correctly!) bet that he would
roll higher on all ten rolls. This seemed implausible to me, but exactly how
implausible? 

In the following we solve the problem exactly, and then confirm our calculations
through simulation.

### Problem

Two players each roll a six-sided, fair die ten times in a row. What is the
probability that Player 1 rolls higher than Player 2 all ten times?

### Analytical solution

Let $\Omega = \{ (x,y) \mid x,y \in \{1, \dots, 6\}\}$ be the sample space, so
that $|\Omega| = 36$. Let $X_1, X_2 \sim \mathcal{U}(1,6)$ be two random
variables representing the rolls of Player 1 and 2 respectively. Let $I$ be an
indicator random variable representing the event that Player 1 wins a roll,
i.e. $P(I) = P(X_1 > X_2)$.

The probability of winning a given roll is straightforward: if Player 1 e.g.
rolls a 6, then he wins if and only if $X_2 \in \{1,2,3,4,5\}$. By the naive
definition of probability, then,

$$
P(I | X_1 = 6) = \frac{|\{1,2,3,4,5\}|}{|\Omega|} = \frac{5}{36} \approx 0.14.
$$

Since the possible events are disjoint, the Law of Total Probability gives us

\begin{align}
P(I) & = P(I | X_1 = 6) + P(I | X_1 = 5) + \dots + P(I | X_1 = 1) \\
     & = \frac{|\{1,2,3,4,5\}|}{|\Omega|} + \dots + \frac{|\{1\}|}{|\Omega|} + \frac{\emptyset}{|\Omega|} \\
     & = \frac{5}{36} + \dots + \frac{1}{36} + \frac{0}{36} \\
     & = \frac{(5)(6)}{(36)(2)} \\
     & = \frac{15}{36} \\
     & = \frac{5}{12} \\
     & \approx 0.42,
\end{align}

where we used the fact that $\sum_{i = 1}^n i = \frac{n(n + 1)}{2}$ in the
fourth equality.

So Player 1 has a 42% chance of winning a given roll.

Now, since all rolls are independent and identically distributed, we can
interpret each roll as a Bernoulli trial with $p = \frac{5}{12}$. Consequently,
if we let $W$ be a random variable representing the total number of rolls won,
see that $W$ is just the sum of $n = 10$ i.i.d. $Bern(p)$ trials, i.e. $W \sim
Binom(10, \frac{5}{12})$. By the definition of the Binomial distribution,

$$
P(W = k) = {n \choose k} p^k(1 - p)^{n-k},
$$

we thus have

\begin{align}
P(W = 10) & = {10 \choose 10} (\frac{5}{12})^{10}(1 - \frac{5}{12})^0 \\
          & = (\frac{5}{12})^{10} \\
          & \approx 0.0001577203.
\end{align}

So the probability of Player 1 winning all ten rolls is about 0.015%.

### Verifying analytical solution through simulation

To make sure we made no mistakes in reasoning or arithmetic in the previous
section, we will verify our calculations with a simple Monte Carlo simulation.

We define a function `roll_dice` which samples with replacement twice from a
uniform distribution, $\mathcal{U}(1,6)$. It then checks if the first sample is
greater than the second and replicates this process `n` times.

```{r roll_dice}
library(tidyverse)
library(scales)
library(ecbutils)

roll_dice <- function(n = 10) {
  replicate(n, {
  roll <- sample(x = 1:6, size = 2, replace = TRUE)
  roll[1] > roll[2]
})
}
```

Next, we specify that we'll run the experiment a million times
(`n_simulations`), and calculate the share of experiments where Player 1 indeed
wins all ten rolls.

```{r simulation}
n_simulations <- 1e6

wins <- replicate(n_simulations, sum(roll_dice()))
table(wins)

cat("\n", "Share of simulations where Player 1 won all ten games:", 
    sum(wins == 10) / n_simulations)
```

Evidently, the simulation results match our earlier calculations to at least
five decimal points, so we're convinced no mistakes were made.

Finally, to confirm that we get the correct results not just for $k = 10$ but
for all $k \in \{0, \dots, 10 \}$, we can plot the distribution of the number of
simulated rolls won by Player 1 and compare this to the probability mass
function of the $Binom(10, \frac{5}{12})$ distribution.

```{r plot}
df_binom <- data_frame(wins = 0:10, 
                       y = dbinom(wins, 10, 15/36) * n_simulations)

ggplot(data_frame(wins), aes(x = wins)) +
  geom_histogram(binwidth = 1, color = "white", 
                 fill = "steelblue", alpha = 0.5) +
  geom_line(aes(y = y), df_binom, color = "red") +
  geom_point(aes(y = y), df_binom, size = 3, color = "red", alpha = 0.5) +
  scale_x_continuous(breaks = pretty_breaks(10), expand = c(0, 0)) +
  scale_y_continuous(breaks = pretty_breaks(10), labels = comma) +
  theme_ecb() +
  labs(x = "\nNumber of rolls won by Player 1")
```
