---
title: "Using a multivariate normal to draw a flower in ggplot2"
author: "EP"
date: "2017-11-24"
categories: 
  - R
tags:
  - ggplot2
editor_options:
  chunk_output_type: console
---

Here's a dumb thing I did over lunch; as I was playing around with the
`geom_density_2d` and random draws from the standard multivariate normal
distribution, I realized that the outcome looked a bit like the flower of a
rose. So I figured I would try to see how difficult it would be to draw the full
rose using `ggplot2`. So here goes.

```{r setup}
knitr::opts_chunk$set(dev = "svg")
```

```{r pkg, message=FALSE, warning=FALSE}
library(tidyverse)
```

We start with the stem. We'll just draw this as a curved line, so `geom_curve`
suits our purposes perfectly.

```{r stem}
p <- ggplot() + 
  coord_equal(1, c(-4, 2), c(-7, 3)) +
  geom_curve(aes(x = -1, y = -7, xend = 0, yend = 0), 
             ncp = 1000, curvature = -0.3, size = 1, 
             color = "olivedrab3")
p
```

Next, we want to draw two leaves on the stem. Since we want to fill these in 
with a color, we'll have to use `geom_polygon`. The solution I came up with is 
very awkward, and wouldn't really generalize to drawing many different types of
leaves, but it's good enough for what we're trying to achieve. (A better
approach might use the `bezier` package.)

```{r geom_leaf}
geom_leaf <- function(x, xend, f, xoffset = 0, yoffset = 0, 
                      xflip = 1, yflip = 1, ...) {
  
  .x <- seq(x, xend, length.out = 100)
  .y <- f(.x)
  
  df <- tibble(x = c(.x, .y), y = c(.y, .x))
  df$x <- xflip * df$x + xoffset
  df$y <- yflip * df$y + yoffset
  
  geom_polygon(aes(x = x, y = y), data = df, ...)
}
```

The primary inputs are start- and end-points on the x-axis, and a function `f` 
to be evaluated over that domain. Together with some parameters for moving and
flipping around the leaves, it then evaluates the function and returns a
`geom_polygon` representing the flower. Here it is in action:

```{r leaf}
f <- function(x) x^2 / 2

p <- p +
  geom_leaf(0, 2, f, -1.6, -4.5, 1, 
            fill = "olivedrab3", color = "palegreen") +
  geom_leaf(0, 2, f, -1.6, -5,  -1, 
            fill = "olivedrab3", color = "palegreen")
p
```

Finally, we create the flower by drawing `n` samples from a standard
multivariate normal distribution and passing these to `stat_density_2d`.

```{r geom_rose}
geom_rose <- function(n, mean = c(0, 0), ...) {
  
  .x <- mvtnorm::rmvnorm(n, mean)
  df <- tibble(x = .x[, 1], y = .x[, 2])
  
  list(
    stat_density_2d(
      aes(x = x, y = y, fill = calc(level)), data = df, 
      geom = "polygon", show.legend = FALSE, color = "grey80"),
    scale_fill_gradient2(...)
  )
}
```

Adding `theme_void` gives us the final result.

```{r rose}
p +
  geom_rose(1000, mean = c(0, 0), 
            low = "red", mid = "purple", high = "pink",
            midpoint = 0.075) +
  theme_void()
```
