---
title: Solving a problem with a generalized outer product
author: EP
date: '2019-08-20'
tags:
  - functional programming
  - math
slug: solving-a-problem-with-a-generalized-outer-product
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  cache = FALSE,
  eval = TRUE,
  fig.width = 4,
  fig.height = 4
)
```

## Introduction

I learned about this interesting problem from a recent [blog
post](https://ryxcommar.com/2019/08/05/a-cool-sql-problem-avoiding-for-loops/)
by [\@ryxcommar](https://twitter.com/ryxcommar).

Imagine you have a time-series of the realized price of some stock:

```{r}
library(ggplot2)

set.seed(2)
df <- data.frame(n = 1:100, y = cumsum(rnorm(100)))
head(df)

p <- ggplot(df, aes(x = n, y = y)) + geom_line()
p
```

With the benefit of hindsight, when would've been the ideal period to buy and
sell that stock? For this stock, inspection of the chart above suggests that the
ideal trade would have been to have bought it in the first period, and sold it
at the peak in period 33:

```{r}
df_trade <- df[df$n %in% c(1, 33), ]

ggplot(df, aes(x = n, y = y)) +
  geom_line() +
  geom_line(data = df_trade, color = "red") +
  geom_point(data = df_trade, color = "red")
```

But how can we solve this problem in general? Also, how would the solution
change if we were forced to hold the stock for $N$ periods before being allowed
to sell it?

In the original blog post, the author has two solutions; a naive solution based
on a for-loop, and a preferred solution based on a series of table joins using
SQL. In this post we will solve the same problem using some functional
programming and a bit of matrix magic.

For kicks, we will solve it using base R for everything except plotting.

## Solution strategy

Let's first state the problem a bit more formally, mostly for notational
purposes.

**Problem**

Let $\overrightarrow{y} = [y_1, \dots, y_n]^T$ be a vector of length $n$ with
real-valued entries. 

Find $i, j \in \mathbb{N}^0$ such that $i - j \geq N$, $N \in
\mathbb{N}$, and $y_i - y_j \geq y_k - y_l$ for all $k, l \in \{1, \dots, n\}$.

**Solution**

Let $M$ be a matrix of pairwise differences of the elements in
$\overrightarrow{y}$:

$$
M = 
\begin{bmatrix}
y_1 - y_1 & y_1 - y_2 & \dots & y_1 - y_n \\
y_2 - y_1 & y_2 - y_2 & \dots & y_2 - y_n \\
\vdots & \vdots & \ddots & \vdots \\
y_n - y_1 &  y_n - y_2 & \dots & y_n - y_n
\end{bmatrix}
=
\begin{bmatrix}
0 & -d_{21} & \dots & -d_{n1} \\
d_{21} & 0 & \dots & -d_{n2} \\
\vdots & \vdots & \ddots & \vdots \\
d_{n1} &  d_{n2} & \dots & 0
\end{bmatrix}.
$$

Now replace all entries of $M$ above the diagonal with $-\infty$:

$$
M' = 
\begin{bmatrix}
0 & -\infty & \dots & -\infty \\
d_{21} & 0 & \dots & -\infty \\
\vdots & \vdots & \ddots & \vdots \\
d_{n1} &  d_{n2} & \dots & 0
\end{bmatrix}.
$$

That is, choosing an "illegal" trade -- where you sell the stock
before you bought it -- results in a payoff of negative infinity.

Let's now assume that $N = 0$, i.e. that that there are no restrictions on for
how long we must hold the stock. With this assumption, and with $M'$
constructed, all we need to do is extract the maximum value $d_{ij}$ from $M'$;
$i$ will be the sell-date, $j$ the buy-date, and $d_{ij}$ the realized profit.

As an aside for the algebra-heads among the readership, replacing illegal
strategies with $-\infty$ works because the extended real number line
$\bar{\mathbb{R}} = \mathbb{R} \cup \{-\infty, \infty\}$ forms a monoid under
the maximum operator with $-\infty$ as the identity element. The proof is left
as an exercise.

Hint: you need to prove 

1. closure: $\max(a, b) \in \mathbb{R} \cup \{-\infty, \infty\}$ for all $a,b
\in \bar{\mathbb{R}}$, 
1. associativity: $\max(\max(a, b), c) = \max(a, \max(b,c))$ for all 
$a,b,c \in \bar{\mathbb{R}}$, 
1. identity element: there exists some unique element $e \in \bar{\mathbb{R}}$ 
such that $\max(a, e) = \max(e, a) = a$ for all $a \in \bar{\mathbb{R}}$.

## Solution in R

With the solution strategy in hand, all we have to do now is implement it in R.

We first create a smaller data set to work with:

```{r}
set.seed(3)
df <- data.frame(n = 1:10, y = cumsum(rnorm(10)))
df

p <- ggplot(df, aes(x = n, y = y)) + geom_line()
p
```

To create the matrix of pairwise differences we can use the generalized outer
product (or "outer difference") operator:

```{r}
m <- outer(df$y, df$y, FUN = `-`)
round(m, 1)
```

Next, we need to replace the upper triangle with `-Inf`:

```{r}
m[upper.tri(m)] <- -Inf
round(m, 1)
```

So with these data, if we buy the stock in the first period and sell it in the
eight period, we'd realize a profit of 0.2 (`m[8, 1]`). The maximum appears in
`m[10, 4]`, indicating that the optimum trade would have been to buy in period 4
and sell in period 10.

To extract the indices of the maximum we *could* first find the max in each
column, and then find in what row that appears:

```{r}
n_buy <- which.max(apply(m, 2, max))
n_sell <- which.max(m[, n_buy])

n_buy
n_sell
```

A nicer way, however, is to rely on the `arr.ind` argument of the `which`
function:

```{r}
which(m == max(m), arr.ind = TRUE, useNames = FALSE)
```

Indeed, let's wrap that into a function that follows our notation: 

```{r}
which.matrix.max <- function(m) {
  idx <- which(m == max(m), arr.ind = TRUE, useNames = FALSE)
  list(i = idx[, 1], j = idx[, 2])
}

strat <- which.matrix.max(m)
strat
```

Finally, we can add a line to the plot of the share price that connects the
buying and selling periods:

```{r}
df_trade <- df[df$n %in% c(strat$j, strat$i), ]

p +
  geom_line(data = df_trade, color = "blue") +
  geom_point(data = df_trade, color = "blue")
```

## Generalizing the solution

What about if $N > 0$? That is, how does our strategy change if we are forced to
hold the stock for $N$ periods before selling it (we're also implicitly assuming
that we must make a trade at some point, i.e. we cannot choose to simply make no
trades)? Well, this simply amounts making the $N$ entries below (and including
the diagonal) infeasible trades, so we simply have to set them to `-Inf` as
well.

If, for example, $N = 7$:

```{r}
N <- 7
m[row(m) - col(m) < N] <- -Inf
round(m, 1)
```

Now the only feasible trades are to buy in one of the first three periods and
sell in the last three periods. The best one would have been to buy in period 2
and sell in period 10.

Since the `m[row(m) - col(m) < N]` solution generalizes the use of `upper.tri`
above, we can wrap this into a function:

```{r}
make_payoff_matrix <- function(x, N) {
  m <- outer(x, x, FUN = `-`)
  m[row(m) - col(m) < N] <- -Inf
  m
}

m <- make_payoff_matrix(df$y, 7)
round(m, 1)
```

We can further combine this with the `which.matrix.max` above to encapsulate the
entire algorithm:

```{r}
get_opt_trade <- function(x, N = 0) {
  
  m <- make_payoff_matrix(x, N)
  idx <- which.matrix.max(m)
  
  data.frame(
    n_buy  = idx$j,
    n_sell = idx$i,
    n_held = idx$i - idx$j,
    profit = x[idx$i] - x[idx$j],
    N      = N
  )
}

get_opt_trade(df$y, 7)
```

Having abstracted away all the complexities of the solution, we can now easily
ask, for example, how the optimum trade and profit changes as a function of the
minimum holding period:

```{r}
# Create a longer time-series
set.seed(11)
n <- 25
df <- data.frame(n = 1:n, y = cumsum(rnorm(n)))

p <- ggplot(df, aes(x = n, y = y)) + geom_line()
p
```

```{r}
df_trade <- lapply(1:10, get_opt_trade, x = df$y)
df_trade <- do.call(rbind, df_trade)

df_trade <- merge(df_trade, df, by.x = "n_buy", by.y = "n")
df_trade <- merge(df_trade, df, by.x = "n_sell", by.y = "n", 
                  suffixes = c("_buy", "_sell"))

p +
  geom_segment(aes(
    x = n_buy, 
    xend = n_sell, 
    y = y_buy, 
    yend = y_sell, 
    color = factor(N)
    ), 
    df_trade) +
  theme(legend.position = "bottom") +
  labs(color = "Minimum \nholding period")
```

The chart shows an interest pattern. The simulated stock here slumped in the
early period, peaked around periods 8-9, then fell almost monotonically
afterwards. So if we faced no or only short minimum holding-limits, the best
trade would have been to buy in the early slump and sell at the peak. However,
if we're forced to hold the stock for 9 or more periods, we would've been unable
to conduct this trade. Our strategy then finds that the best trade (which
minimizes losses) would have been to have bought and sold once the fall in stock
prices had levelled off towards the end of the period.
