# ---
# title: Implementing a perfect-playing Nim opponent in R
# author: EP
# date: '2017-07-30'
# categories:
#   - R
# tags:
#   - functional programming
#   - combinatorial game theory
# ---

## It's time for some (combinatorial) game theory

In a previous post we wrote a program that allows you to play a game of Nim
against another human opponent. Now, to be perfectly frank, Nim is not a very
fun game to play, so I think many readers would be hard-pressed to find somebody
to play against. Luckily, it turns out that it's very simple to modify our
initial program to also allow for play against a computer-opponent. To
understand how this is possible, however, we must first discuss a bit of
algebra.

As mentioned in the previous post, the game of Nim gives rise to a special type
of numbers called, naturally, "nimbers" (mathematicians are clearly not immune
to dad jokes). These numbers look similar to regular integers, but they are
added together by performing a bitwise xor operation on their binary
representations. So how does this work?

Consider the numbers 4 and 7. In binary, these would be represented as 100 and
111, respectively. The xor operator takes two bits (0s or 1s) as input and
returns a 1 if exactly one of the two inputs are 1, and otherwise returns 0. For
example:

```{r}
xor(0,0)
xor(0,1)
xor(1,0)
xor(1,1)
```

To perform the bitwise xor operation on two numbers, therefore, means we
take two numbers, convert them to binary, line them up, and perform the xor
operation on each consequtive pair of bits. In the case of 4 and 7 we thus get:

```{r bits}
head(intToBits(4))
head(intToBits(7))
head(intToBits(bitwXor(4, 7)))
```

(R gives the binary representation from left to right, so e.g. 4 = 001 here.
We're also wrapping everything in `head` to avoid printing out a ton of zeros).
So the `bitwXor` first looks at 0 and 1 and evaluates to 1, then 1 and 0 gives 1
again, and finally 1 and 1 gives 0. So `bitwXor(4,7)` = 110 (in binary) = 3 (in
decimal).

We said earlier that bitwise xor is how you add together two nimbers, so it
would be convenient if we could use standard notation and just write $4 + 7 =
3$. Luckily, R's beautifully minimalist S3 system is perfectly suited for this.

We start by defining a constructor function, which simply tags a number as a
nimber:

```{r nimbers}
nimber <- function(x) {
  structure(x, class = c("nimber", class(x)))
}

nimber(5)
```

We then implement a pretty-printing method (the `unclass` call here is necessary
to avoid infinite recursive calls to `print`.):

```{r}
print.nimber <- function(x) {
  cat("Nimbers:\n")
  print(unclass(x))
}

nimber(1:5)
```

With that out of the way, we can now overload the `+` operator so that nimber
addition is defined as bitwise xor:

```{r}
`+.nimber` <- function(e1, e2) {
  nimber(bitwXor(e1, e2))
}

nimber(4) + nimber(7)
```

We can also implement a `[` method, so that we can index into a vector of
nimbers without those nimbers being converted into numbers:

```{r}
nimber(1:5)[3]

`[.nimber` <- function(x, i) {
  nimber(NextMethod())
}

nimber(1:5)[3]
```

Finally, we would like to be able to take sum of a vector of nimbers. We implement the `sum` method using `purrr::reduce`:

```{r}
library(purrr)

sum.nimber <- function(x, na.rm = TRUE) {
  reduce(x, `+`, .init = nimber(0))
}

sum(nimber(c(4, 7)))
sum(nimber(1:5))
```

(The mathematically-inclined reader might wonder if this is a legitimate way to
sum a set of nimbers. Indeed it is; bitwise xor is associative with 0 as the
identity element. In fact, every nimber is its own inverse (e.g. 100101 bitwised
xor'ed with itself will produce 000000 = 0, the identity element) and the order
in which we perform the operation is irrelevant. Hence, the nimbers actually
form an abelian group under nim-addition. To be even more precise, since every
element except the identity is of order 2, they form a
(https://proofwiki.org/wiki/Definition%3aBoolean_Group)[Boolean group])

So what does all of this have to do with a computer playing Nim? Well, it turns
out that if we treat the number of pebbles in a Nim heap as a nimber, then any
game board can be represented by nim-summing up all the heaps. This sum will
then either be zero or non-zero, and a player can always make a move that takes
a non-zero-sum board and produces a zero-sum board. The key insight from the
mathematical analysis of Nim is that the optimal move is always to produce a
zero-sum board. (We omit all proofs and details here. See the
(https://en.wikipedia.org/wiki/Nim#Mathematical_theory)[Wikipedia entry] for
some quite accessible proofs and examples).

So we can define an `optimal_move` function, which takes a board as input,
calculates the nim-sum of the board, randomly selects a row among the rows that
have a certain desired property, removes the number of pebbles necessary to
produce a zero-sum board, and returns the optimal move:

```{r optimal_move}
resample <- function(x, ...) {
  x[sample.int(length(x), ...)]
}

optimal_move <- function(board) {
  board_sum <- sum(board)
  row <- resample(which(board + board_sum < board), 1)
  num <- board[row] - (board[row] + board_sum)
  c("row" = row, "num" = num)
}
```

(`resample` is just a more stable version of `sample`). Below, we make sure that
the board is always of class `nimber`, so that the `sum` call gets dispatched to
`sum.nimber`.

While we're at it, we can also implement a random move function:

```{r}
random_move <- function(board) {
  row <- resample(seq_along(board[board != 0]), 1)
  num <- resample(range(1, board[row]), 1)
  c("row" = row, "num" = num)
}
```

```{r}
computer_move <- function(board) {
  board_sum <- sum(board)
  when(board_sum,
       . == 0 ~ random_move(board),
       . != 0 ~ optimal_move(board))
}
```

Thanks to the way we originally structured the game, refactoring the `nim`
function to allow for a computer opponent is straightforward. We just have to
add a `comp_opponent` argument, make sure that the board is always of class
`nimber`, and generate a computer move if it's the computer's turn to move:

```{r game_w_computer_moves}
# These remain unchanged
next_player <- function(player) if(player == 1) 2 else 1
valid_move  <- function(board, row, num) num > 0 && board[row] >= num
finished    <- function(board) all(board == 0)
update      <- function(board, row, num) `[<-`(board, row, board[row] - num)
disp_row    <- function(row, num) cat(row, ":", rep("*", num), "\n")
disp_board  <- function(board) walk2(seq_along(board), board, disp_row)
get_move    <- function() {
  row <- as.integer(readline("Choose a row: "))
  num <- as.integer(readline("Number of stars to remove: "))
  c("row" = row, "num" = num)
}

nim <- function(board = 5:1, player = 1, comp_opponent = TRUE) {
  if(!inherits(board, "nimber")) {
    board <- nimber(board)
  }

  if(finished(board)) {
    cat("Game over. Player", next_player(player), "wins!")
    return(invisible())
  }

  disp_board(board)
  cat("Player", player, "\n")
  play <- if(player == 2 && comp_opponent)
    computer_move(board)
  else get_move()

  row  <- play["row"]
  num  <- play["num"]

  if(!valid_move(board, row, num)) {
    cat("Not a valid move. Try again.\n")
    nim(board, player)
  } else nim(update(board, row, num), next_player(player))
}
```
